On the spatial aggregation of ecosystem condition indicators

### 2025-02-05 manuscript/manuscript.qmd
Callouts as boxes

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
  - upscaling
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  cache: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

@@

Similar to variables, the reference levels used for normalising variables also have a spatial assignment, which can be different from that of the variable.
For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g. a natural zero).
Variables should not be normalised based on reference levels unless they are at a commensurable spatial scale.
Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the reference condition, then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition, simply due to natural variation in where the animals are at any give time.
Averaging (arguably the most common spatial aggregation function) the indicator values across sub-regions would not produce a regional indicator value that shows the region as being perfectly aligned with the reference condition, since truncation of variable values above X~100~ (assuming X~100~ \> X~0~) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
See @nte-billberry for another example.

::: {#nte-billberry .callout-note appearance="minimal" icon="false"}
### Billberry example

@@

Complete, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
This last type of data is very common.
A lot of demographic information for example is made available at the scale of administrative units, rather than at the individual level (which would be very ineffective and problematic also in terms of protecting sensitive information).
Similar pre-aggregation is, however, ambiguous and true for most complete datasets.
Consider as an example data from LiDAR (Light Detection And Ranging).
The raw data is a point cloud, yet the most commonly available datasets are rasters, such as surface or canopy models; wall-to-wall generalisations of the point cloud.
Also data such as satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (@nte-rel).
This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
This will clearly affect the suitability of such data for identifying areas of high risk (very degraded areas, for example).
The problem is the same when we deliberately aggregate data later in the indicator development process, and we always need to ensure that the spatial aggregation function we use does not have any unintentional effect on the type of inference that we can draw from the metric.

For sparse data we are considering data where variable values are assigned to singular points or small areas.
In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

::: {#nte-rel .callout-note appearance="minimal" icon="false"}
### Related concepts

##### Vertical aggregation {.unnumbered}

@@

##### Upscaling {.unnumbered}

@@

##### Modifiable area unit problem {.unnumbered}

@@

##### Ecological fallacy {.unnumbered}

@@

The spatial aggregation of data can lead to biases arising from a number of sources.
For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
In this paper, however, we are mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
They showed that when we take the average of two variables, each of which have associated attribute values (e.g. indicator values), then the attribute of the average variable value differs from the average of the attributes of the original variables.
This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.
In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different **aggregation pathways** during indicator development.

@@

Pathways 7-8 describe the spatial aggregation of multiple indicators, as they all start with the normalisation step.
These are not the main topic of this paper, but are included here for completeness.
In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
In pathway 8, spatial aggregation is only done on individual indicators.
The two pathways differ in how they handle more complicated schemes for weighted averages, with pathway 8 being more flexible, but it may interfere with the principle of commutativity [@unitednations2021].

>>>>

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
  - upscaling
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  cache: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
callout-appearance: simple
callout-icon: false
---

@@

Similar to variables, the reference levels used for normalising variables also have a spatial assignment, which can be different from that of the variable.
For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g. a natural zero).
Variables should not be normalised based on reference levels unless they are at a commensurable spatial scale.
Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the reference condition, then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition, simply due to natural variation in where the animals are at any give time.
Averaging (arguably the most common spatial aggregation function) the indicator values across sub-regions would not produce a regional indicator value that shows the region as being perfectly aligned with the reference condition, since truncation of variable values above X~100~ (assuming X~100~ \> X~0~) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
See Box 1 for another example.

::: {.callout-note}

**Box 1 -- Billberry example** \

@@

Complete, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
This last type of data is very common.
A lot of demographic information for example is made available at the scale of administrative units, rather than at the individual level (which would be very ineffective and problematic also in terms of protecting sensitive information).
Similar pre-aggregation is, however, ambiguous and true for most complete datasets.
Consider as an example data from LiDAR (Light Detection And Ranging).
The raw data is a point cloud, yet the most commonly available datasets are rasters, such as surface or canopy models; wall-to-wall generalisations of the point cloud.
Also data such as satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (Box 2).
This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
This will clearly affect the suitability of such data for identifying areas of high risk (very degraded areas, for example).
The problem is the same when we deliberately aggregate data later in the indicator development process, and we always need to ensure that the spatial aggregation function we use does not have any unintentional effect on the type of inference that we can draw from the metric.

For sparse data we are considering data where variable values are assigned to singular points or small areas.
In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; Box 2).
It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

::: {.callout-note}

**Box 2 -- Related concepts** \

**Vertical aggregation**

@@

**Upscaling**

@@

**Modifiable area unit problem**

@@

**Ecological fallacy**

@@

The spatial aggregation of data can lead to biases arising from a number of sources.
For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
In this paper, however, we are mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
They showed that when we take the average of two variables, each of which have associated attribute values (e.g. indicator values), then the attribute of the average variable value differs from the average of the attributes of the original variables.
This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.
In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different **aggregation pathways** during indicator development.

@@

Pathways 7-8 describe the spatial aggregation of multiple indicators, as they all start with the normalisation step.
These are not the main topic of this paper, but are included here for completeness.
In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
In pathway 8, spatial aggregation is only done on individual indicators.
The two pathways differ in how they handle more complicated schemes for weighted averages, with pathway 8 being more flexible, but it may interfere with the principle of commutativity [@unitednations2024].
